---
title: "Two-sample T-test (BIOEST2226)"
output: html_document
date: "2022-11-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

For my bio statistics class on UNM they gave us a random data set to each student. The goal was to create a _fictional study_ and apply the right hypothesis testing. 

My data set is composed by two variables, one quantitative and one categorical with two categories.

Originally, I made this analysis using [Infostat](https://www.infostat.com.ar/), as we were though how to use the software on the course.

## The study

Among calf raisers there is a strong preference for calving time. The vast majority favor winter births so that the weaning period coincides with the beginning of spring. 
The reasons given by breeders are diverse, one of them being that calves born in winter reach a higher weight at weaning than those born in summer, which makes production at this time more attractive.


Similarly, we would like to know if the same can be concluded for the breeding of foals to be used in high performance sports. In this case, having smaller foals could generate an advantage when selecting them for sports in which agility is an important factor. 

For this purpose, we collected data from 400 births in Quarter Horse farms in Argentina, for the period 2020 - 2021. 
The data have been tabulated detailing the weight of the foal at birth. Each foal was assigned a group, depending on the season of the year in which they were born. The sample consisted of two groups, with 210 foals born in summer (grupo 1) and 190 in winter (grupo 2).

_Please have in mind all this context is fictional._

## Loading data

```{r message=FALSE}
library(dplyr)
library(ggplot2)
library(cowplot) # to place two ggplots side-by-side
```

```{r}
data <- read.csv("data/datos21.txt", header = FALSE, sep = "\t", col.names = c('weight', 'group'))

```

## Descriptive statictics

First of all, we'll get a summary for each sample.

```{r}
data %>% group_by(group) %>% 
  summarise(min = min(weight),
            max = max(weight),
            mean = mean(weight),
            sd = sd(weight),
            n = n(),
            q1 = quantile(weight, .25),
            q3 = quantile(weight, .75)) 
```

It can be observed that the mean weight in grupo2 (winter) is 10 kg higher than for grupo1 (summer). 
For both samples the median value is close to the mean, suggesting that the data are normally distributed. 
The standard deviations are very similar for both samples, indicating that they differ similarly from their mean.

We could observe the same using a boxplot.

```{r}
data %>% ggplot(aes(x=group, y=weight, fill=group)) + 
  geom_boxplot() + 
  scale_fill_manual(values = c("#e74c3c", "#2ecc71")) + 
  labs(
    title = "Births, 2020 - 2021 period",
    x = "Group",
    y = "Weight (kg)"
  )
```

Here we can also notice more grafically that extreme minimum weights were recorded in summer. In addition, it is evident that the minimum values of the group of foals born in winter overlaps with the maximum values of those born in summer.



```{r}
data %>%
  ggplot( aes(x=weight, color=group, fill=group)) +
    geom_histogram(aes(y=after_stat(density)),bins=16, colour = 1) +
    scale_fill_manual(values = c("#e74c3c", "#2ecc71")) + 
    xlab("Group") +
    ylab("Frecuency of births") +
    facet_wrap(~group)

```

The histograms allow us to see at a glance the distribution of the data.

The left one indicates that the sample of foals born in summer has no skew. 
The right one, for foals born in winter, has a negative skew. 


A priori, we could tell there is a tendency to obtain smaller foals if they are born at higher temperatures.

## Hypothesis

Null: No differences are observed in the average weight of foals born at different times of the year. Those born in summer weigh the same as those born in winter.

$\mu_{S} - \mu_{W} = 0$ 

Alternative: Differences are observed in the average weight of foals born at different times of the year, with those born in summer being less heavy than those born in winter.

$\mu_{S} - \mu_{W} < 0$ 



Being

$\mu_{S}$ : Mean weight in kilograms of foals born in summer

$\mu_{W}$ : Mean weight in kilograms of foals born in winter

## Hypothesis test

```{r}
summer <- filter(data, group == "grupo1") %>% select(weight) %>% unlist
winter <- filter(data, group == "grupo2") %>% select(weight) %>% unlist
```

### Assumptions

#### Normality

```{r}
data %>% ggplot(aes(sample = weight)) +
  stat_qq() +
  stat_qq_line() +
  xlab("Theoretical") +
  ylab("Observed") +
  facet_wrap(~group)
```

#### Equal variances



### Results

```{r}
t.test(
  x = summer, 
  y = winter, 
  alternative="less", 
  mu = 0, 
  paired=FALSE, 
  var.equal=TRUE, 
  conf.level=0.95
  )
```